RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-08-29 formalframework.tex
added some comments and minor things

>>>>

\todo{M}{F,M}{\url{https://en.wikipedia.org/wiki/Atomic_formula}
I think atom just means there are no connectives or negations. I used this terminology since it is used in logic programming which is very concrete and clearly defined in terms of syntax. Predicates have such a wide meaning that it may confuse the reader: \url{https://en.wikipedia.org/wiki/Predicate_(mathematical_logic)}. Anyway I am also not very happy with these definitions yet. I'll give it another try taking your suggestions into account.}

### 2017-08-29 introduction.tex
added some comments and minor things

A further problem is that in goal models in general the rationale behind any modeling decision is static, that is, current goal modeling languages have limited support for reasoning about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what its effect is on the goal model. Similarly, with existing goal modelling languages one can change a part of the goal model, but it is not possible to reason about whether or not this new goal model is consistent with the underlying beliefs and arguments. This becomes even more problematic if the stakeholders constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs of another group of stakeholders. The disconnect between the goal models and their underlying beliefs and opinions may further lead to a poor understanding of the problem and solution domain, which is an important reason of RE project failure~\cite{curtis1988field}.

To summarize, what is needed is a way of recording the rationales (beliefs, opinions, discussions, ideas) underlying a goal model. It should be possible to see how these rationales changed during the goal modeling process, and the rationales hould be clearly linked to the various elements of the resulting goal model. In order to be able to do this, we propose a framework with tool-support that combines traditional goal modeling approaches with argumentation techniques from Artificial Intelligence (AI) research~\cite{BenchCaponDunne2007}. We have identified \textbf{five important requirements} for our framework:

\begin{enumerate}
\item 
The argumentation techniques should be close to the actual discussions of stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the goal model on the underlying arguments, and vice versa.\todo{F}{M,S}{Given the current logical framework, is it possible to compute effect of changes in GM on ARG? It seems to me that we can only do the other way around with the algorithms.}
\item 
There should be a methodology for the framework to guide the practitioners in its application in real cases.
\item 
The framework must have software tool support.
\end{enumerate}

@@

To answer the research question above and solve some of the concerns mentioned in Subsection~\ref{sect:problem_statement}, we propose a framework called RationalGRL. The RationalGRL framework combines the Goal-oriented Requirements Language (GRL) with a technique from argumentation theory and discourse modeling called \emph{argument schemes}~\cite{walton-etal2004}. \todo{Floris}{Floris}{mention Singh et al. argument schemes in RE here.}

>>>>

A further problem is that the rationale behind goal modeling decisions is usually static, that is, current goal modeling languages have limited support for reasoning about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what its effect is on the goal model. Similarly, with existing goal modelling languages one can change a part of the goal model, but it is not possible to reason about whether or not this new goal model is consistent with the underlying beliefs and arguments. This becomes even more problematic if the stakeholders constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs of another group of stakeholders. The disconnect between the goal models and their underlying beliefs and opinions may further lead to a poor understanding of the problem and solution domain, which is an important reason of RE project failure~\cite{curtis1988field}.

To summarize, what is needed is a way of recording the rationales (beliefs, opinions, discussions, ideas) underlying a goal model. It should be possible to see how these rationales changed during the goal modeling process, and the rationales should be clearly linked to the various elements of the resulting goal model. In order to be able to do this, we propose a framework with tool-support that combines traditional goal modeling approaches with argumentation techniques from Artificial Intelligence (AI) research~\cite{BenchCaponDunne2007}. We have identified \textbf{five important requirements} for our framework:

\begin{enumerate}
\item 
The argumentation techniques should be close to the actual discussions of stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the goal model on the underlying arguments, and vice versa.
\todo{F}{M,S}{Given the current logical framework, is it possible to compute effect of changes in GM on ARG? It seems to me that we can only do the other way around with the algorithms. (M): In a sense this is possible. In our formalization, each GRL element and link can be translated into an instantiation of an argument scheme, so adding GRL element simply means adding more instantiate argument schemes, who can then be attack by critical questions. But I think it is a good idea to make this more explicit.} 
\item 
There should be a methodology for the framework to guide the practitioners in its application in real cases.
\item 
The framework must have software tool support.
\end{enumerate}

@@

To answer the research question above and solve some of the concerns mentioned in Subsection~\ref{sect:problem_statement}, we propose a framework called RationalGRL. The RationalGRL framework combines the Goal-oriented Requirements Language (GRL) with a technique from argumentation theory and discourse modeling called \emph{argument schemes}~\cite{walton-etal2004}. \todo{Floris}{Floris}{mention Singh et al. argument schemes in RE here. (Marc): I wouldn't go into much detail here. Perhaps one sentence and then refer to the related work section.}
